Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2017-12-21 PAPER.tex
edited affiliations
edited gianni's affilication
made gianni edits

\author{
Carina Cheng\altaffilmark{1},
Aaron R. Parsons\altaffilmark{1,2},
Matthew Kolopanis \altaffilmark{3},
Daniel C. Jacobs\altaffilmark{3}, 
Saul A. Kohn\altaffilmark{4},
Adrian Liu\altaffilmark{1,5}, 
James E.~Aguirre\altaffilmark{4},
Zaki S. Ali\altaffilmark{1}, 
Gianni Bernardi\altaffilmark{6,7,8}, 
Richard F. Bradley\altaffilmark{9,10,11},
Chris L. Carilli\altaffilmark{12,13},
David R. DeBoer\altaffilmark{2}, 
Matthew R. Dexter\altaffilmark{2},
Joshua S. Dillon\altaffilmark{1},
Pat Klima\altaffilmark{7},
David H. E. MacMahon\altaffilmark{2},
David F. Moore\altaffilmark{4},
Chuneeta D. Nunhokee\altaffilmark{10},
Jonathan C. Pober\altaffilmark{14}, 
William P. Walbrugh\altaffilmark{9},
Andre Walker\altaffilmark{9}
}

\altaffiltext{1}{Astronomy Dept., U. California, Berkeley, CA}
\altaffiltext{2}{Radio Astronomy Lab., U. California, Berkeley CA}
\altaffiltext{3}{School of Earth and Space Exploration, Arizona State U., Tempe AZ}
\altaffiltext{4}{Dept. of Physics and Astronomy, U. Penn., Philadelphia PA} 
\altaffiltext{5}{Berkeley Center for Cosmological Physics, Berkeley, CA} 
\altaffiltext{6}{Square Kilometer Array, S. Africa, Cape Town South Africa}
\altaffiltext{7}{Dept. of Physics and Electronics, Rhodes University, South Africa}
\altaffiltext{8}{Harvard-Smithsonian Cen. for Astrophysics, Cambridge MA}
\altaffiltext{9}{Dept. of Electrical and Computer Engineering, U. Virginia, Charlottesville VA}
\altaffiltext{10}{National Radio Astronomy Obs., Charlottesville VA}
\altaffiltext{11}{Dept. of Astronomy, U. Virginia, Charlottesville VA}
\altaffiltext{12}{National Radio Astronomy Obs., Socorro NM}
\altaffiltext{13}{Cavendish Lab., Cambridge UK}
\altaffiltext{14}{Dept. of Physics, Brown University, Providence RI}

@@

Although a detection of the EoR remains elusive, there are several radio telescope experiments that have succeeded in using 
the 21 cm signal from hydrogen to place constraints on the brightness of the signal. Examples of experiments investigating the 
mean brightness temperature of the EoR relative to the CMB are the Experiment to Detect the Global EoR Signature (EDGES; 
\citealt{bowman2010}), the Large Aperture Experiment to Detect the Dark Ages (LEDA; \citealt{greenhill_bernardi2012}), the 
Dark Ages Radio Explorer (DARE; \citealt{burns2012}), the Sonda Cosmol\'ogica de las Islas para la Detecci\'on de 
Hidr\'ogeno NeutroSciHi (SCI-HI; \citealt{voytek2014}), the Broadband Instrument for Global HydrOgen ReioNisation Signal 
(BIGHORNS; \citealt{sokolowski2015}), and the Shaped Antenna measurement of the background RAdio Spectrum (SARAS; 
\citealt{patra2015}). Radio interferometers which seek to measure statistical power spectra include the Giant Metre-wave 
Radio Telescope (GMRT; \citealt{paciga_et_al2013}), the LOw Frequency ARray (LOFAR; \citealt{van_haarlem_et_al2013}), 
the Murchison Widefield Array (MWA; \citealt{tingay_et_al2013}), the 21 Centimeter Array (21CMA; 
\citealt{peterson_et_al2004}; \citealt{wu2009}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of 
Reionization Array (HERA), which is currently being built, is a next-generation instrument that aims to combine lessons 
learned from previous experiments and is forecasted to be able to make a high-significance power spectrum 
detection with an eventual $350$ elements using current analysis techniques (\citealt{pober_et_al2014}; \citealt{liu_parsons_2016}; \citealt{deboer_et_al2017}).

@@

\begin{equation}
\label{eq:qhat}
\hat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}.
\end{equation}
Here, \noindent $\textbf{Q}$ is a family of matrices that takes our frequency-domain visibilities and Fourier-transforms them into 
power spectrum space, while also taking into account physical constants such as the Boltzmann constant used to convert Jy to 
Kelvin, as well as cosmological scalings. The index $\alpha$ denotes a waveband in $k_{\parallel}$, where $k_{\parallel}$ is the 
Fourier-dual to frequency under the delay approximation (\citealt{parsons_et_al2012b}). $\textbf{R}$ is a weighting matrix --- as 
an example, inverse covariance weighting would set $\textbf{R} = \textbf{C}^{-1}$ and an unweighted case would use $
\textbf{R} = \textbf{I}$, the identity matrix.

@@

As a first test, we model the covariance matrix of EoR as a proof of concept that if perfect models are known, signal loss can be 
avoided. We know that our simulated EoR signal should have a covariance matrix that mimics the identity matrix, with its 
variance encoded along the diagonal. We model $\textbf{C}_{EoR}$ as such, instead of computing it based on $\textbf{x}
_{EoR}$ itself. Next, we add $\textbf{C}_{EoR} + \hat{\textbf{C}}_{FG}$ (where $\hat{\textbf{C}}_{FG} = \langle\textbf{x}_{FG}
\textbf{x}_{FG}^{\dagger}\rangle_{t}$) to obtain a final $\hat{\textbf{C}}$ to use in weighting. In Figure \ref{fig:toy_sigloss8} (upper 
left), we see that there is negligible signal loss. This is because by modeling $\textbf{C}_{EoR}$, we avoid over-fitting EoR fluctuations in the data that our model doesn't know about (but an empirically derived $\hat{\textbf{C}}$ would). This is also shown by comparing the (steeper) green and (flatter) red curves in Figure \ref{fig:toy_sigloss14}. We don't see perfect EoR recovery though, as the 
green and red curves vary slightly in Figure \ref{fig:toy_sigloss8}. This deviation is a consequence of the difference between the 
true and modeled covariance matrices. In practice such a weighting option is not feasible, as we do not know $\textbf{C}_{EoR}
$.

@@

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} flattens out the eigenspectrum of $\hat{\textbf{C}}$ a different way - 
by only using the first three eigenmodes. Recalling that the foregrounds can be described entirely by the first eigenmode, this 
method intentionally projects out modes that are EoR-dominated by replacing all but the three highest weights in the 
eigenspectrum with $1$'s (equal weights). Again, flattening the eigenspectrum results in negligible signal loss. However, we do 
not perfectly recover the shape of EoR because we lost information when projecting out certain modes.

>>>>

\author{
Carina Cheng\altaffilmark{1},
Aaron R. Parsons\altaffilmark{1,2},
Matthew Kolopanis \altaffilmark{3},
Daniel C. Jacobs\altaffilmark{3}, 
Saul A. Kohn\altaffilmark{4},
Adrian Liu\altaffilmark{1,5}, 
James E.~Aguirre\altaffilmark{4},
Zaki S. Ali\altaffilmark{1}, 
Gianni Bernardi\altaffilmark{6,7,8}, 
Richard F. Bradley\altaffilmark{9,10,11},
Chris L. Carilli\altaffilmark{12,13},
David R. DeBoer\altaffilmark{2}, 
Matthew R. Dexter\altaffilmark{2},
Joshua S. Dillon\altaffilmark{1},
Pat Klima\altaffilmark{10},
David H. E. MacMahon\altaffilmark{2},
David F. Moore\altaffilmark{4},
Chuneeta D. Nunhokee\altaffilmark{7},
Jonathan C. Pober\altaffilmark{14}, 
William P. Walbrugh\altaffilmark{6},
Andre Walker\altaffilmark{6}
}

\altaffiltext{1}{Astronomy Dept., U. California, Berkeley, CA}
\altaffiltext{2}{Radio Astronomy Lab., U. California, Berkeley CA}
\altaffiltext{3}{School of Earth and Space Exploration, Arizona State U., Tempe AZ}
\altaffiltext{4}{Dept. of Physics and Astronomy, U. Penn., Philadelphia PA} 
\altaffiltext{5}{Berkeley Center for Cosmological Physics, Berkeley, CA} 
\altaffiltext{6}{Square Kilometer Array, S. Africa, Cape Town South Africa}
\altaffiltext{7}{Dept. of Physics and Electronics, Rhodes University, South Africa}
\altaffiltext{8}{INAF-Instituto di Radioastronomia, Bologna Italy}
\altaffiltext{9}{Dept. of Electrical and Computer Engineering, U. Virginia, Charlottesville VA}
\altaffiltext{10}{National Radio Astronomy Obs., Charlottesville VA}
\altaffiltext{11}{Dept. of Astronomy, U. Virginia, Charlottesville VA}
\altaffiltext{12}{National Radio Astronomy Obs., Socorro NM}
\altaffiltext{13}{Cavendish Lab., Cambridge UK}
\altaffiltext{14}{Dept. of Physics, Brown University, Providence RI}

@@

Although a detection of the EoR remains elusive, there are several radio telescope experiments that have succeeded in using 
the 21 cm signal from hydrogen to place constraints on the brightness of the signal. Examples of experiments investigating the 
mean brightness temperature of the EoR relative to the CMB are the Experiment to Detect the Global EoR Signature (EDGES; 
\citealt{bowman2010}), the Large Aperture Experiment to Detect the Dark Ages (LEDA; \citealt{bernardi_et_al2016}), the 
Dark Ages Radio Explorer (DARE; \citealt{burns2012}), the Sonda Cosmol\'ogica de las Islas para la Detecci\'on de 
Hidr\'ogeno NeutroSciHi (SCI-HI; \citealt{voytek2014}), the Broadband Instrument for Global HydrOgen ReioNisation Signal 
(BIGHORNS; \citealt{sokolowski2015}), and the Shaped Antenna measurement of the background RAdio Spectrum (SARAS; 
\citealt{patra2015}). Radio interferometers which seek to measure statistical power spectra include the Giant Metre-wave 
Radio Telescope (GMRT; \citealt{paciga_et_al2013}), the LOw Frequency ARray (LOFAR; \citealt{van_haarlem_et_al2013}), 
the Murchison Widefield Array (MWA; \citealt{tingay_et_al2013}), the 21 Centimeter Array (21CMA; 
\citealt{peterson_et_al2004}; \citealt{wu2009}), and PAPER (\citealt{parsons_et_al2010}). The Hydrogen Epoch of 
Reionization Array (HERA), which is currently being built, is a next-generation instrument that aims to combine lessons 
learned from previous experiments and is forecasted to be able to make a high-significance power spectrum 
detection with an eventual $350$ elements using current analysis techniques (\citealt{pober_et_al2014}; \citealt{liu_parsons_2016}; \citealt{deboer_et_al2017}).

@@

\begin{equation}
\label{eq:qhat}
\hat{q}_{\alpha} = \frac{1}{2}\textbf{x}^{\dagger}\textbf{R}\textbf{Q}^{\alpha}\textbf{R}\textbf{x}.
\end{equation}
Here, \noindent $\textbf{Q}$ is a family of matrices that takes our frequency-domain visibilities and Fourier-transforms them into 
power spectrum space, while also taking into account physical constants such as the Boltzmann constant used to convert Jy to 
Kelvin, as well as cosmological scalings. It is evaluated as $\textbf{Q}^{\alpha} \equiv \frac{\delta\textbf{C}}{\delta p_{\alpha}}$, or the derivative of the covariance with respect to the true bandpower in $\alpha$, where $\alpha$ denotes a waveband in $k_{\parallel}$ ($k_{\parallel}$ is the 
Fourier-dual to frequency under the delay approximation (\citealt{parsons_et_al2012b})). $\textbf{R}$ is a weighting matrix --- as 
an example, inverse covariance weighting would set $\textbf{R} = \textbf{C}^{-1}$ and an unweighted case would use $
\textbf{R} = \textbf{I}$, the identity matrix.

@@

As a first test, we model the covariance matrix of EoR as a proof of concept that if perfect models are known, signal loss can be 
avoided. We know that our simulated EoR signal should have a covariance matrix that mimics the identity matrix, with its 
variance encoded along the diagonal. We model $\textbf{C}_{EoR}$ as such (i.e. the identity), instead of computing it based on $\textbf{x}
_{EoR}$ itself. Next, we add $\textbf{C}_{EoR} + \hat{\textbf{C}}_{FG}$ (where $\hat{\textbf{C}}_{FG} = \langle\textbf{x}_{FG}
\textbf{x}_{FG}^{\dagger}\rangle_{t}$) to obtain a final $\hat{\textbf{C}}$ to use in weighting. In Figure \ref{fig:toy_sigloss8} (upper 
left), we see that there is negligible signal loss. This is because by modeling $\textbf{C}_{EoR}$, we avoid over-fitting EoR fluctuations in the data that our model doesn't know about (but an empirically derived $\hat{\textbf{C}}$ would). This is also shown by comparing the (steeper) green and (flatter) red curves in Figure \ref{fig:toy_sigloss14}. We don't see perfect EoR recovery though, as the 
green and red curves vary slightly in Figure \ref{fig:toy_sigloss8}. This deviation is a consequence of the difference between the 
true and modeled covariance matrices. In practice such a weighting option is not feasible, as we do not know $\textbf{C}_{EoR}
$.

@@

The third panel (bottom left) in Figure \ref{fig:toy_sigloss8} flattens out the eigenspectrum of $\hat{\textbf{C}}$ a different way - 
by only using the first three eigenmodes. Recalling that our toy model foregrounds can be described entirely by the first eigenmode, this 
method intentionally projects out modes that are EoR-dominated by replacing all but the three highest weights in the 
eigenspectrum with $1$'s (equal weights). Again, flattening the eigenspectrum results in negligible signal loss. However, we do 
not perfectly recover the shape of EoR because we lost information when projecting out certain modes. While this case is illuminating for the toy model, in practice it is not obvious which eigenmodes are FG vs. EoR dominated, so selecting a subset of modes to down-weight is not trivial.
